Reconstructing Propositional Proofs in Type Theory

### 2017-11-23 paper/sections/introduction.tex
First version of the document - November 22, 2017 (#51)
* [ paper ] two first paragraphs in the introduction.

* [ bib] Fixed warning.

(Warning--empty publisher in Bertot2004).

* [ bib ] Fixed warning.

(Warning--empty booktitle in Bove2005).

* [ bib ] Fixed warning.

(Warning--empty booktitle in Een2004).

* [ bib ] Fixed warning.

(Warning--empty journal in Altenkirch2015).

* [ bib ] Fixed warning.

(Warning--empty year in deNivelle2003).

* [ paper ] add the third paragraph in the introduction.

* [ paper ] added the fourth paragraph in the introduction.

* [ paper
] fixed some text in

In order to give confidence to the \ATP users, many of these systems
have started to include in their outputs the full derivations
associated to the proved theorems. However, existing research
recognizes that in many cases these proofs encode non-trivial
reasoning hard to reconstruct and therefore hard to
verify~\cite{paulson2007source,Keller2013}.

Proof-reconstruction addresses this problem. It becomes in mostly
cases a reverse engineering task to verify the prover reasoning. That
usually happens when the reconstruction is made by another and not by
the developers of the \ATP. The presentation of the proofs generated
by the prover plays an important role in proof-reconstruction. To
verify such a proof, it is convenient to have it in a consistent
format, that is, a full script describing the derivation step-by-step
with  exhaustive details and without ambiguities. For example, for
classical propositional logic (henceforth \CPL) from a list of at
least forty \ATPs---available from the Web service
\name{SystemOnTPTP} of the TPTP World\footnote{\url{http://www.cs.miami.edu/~tptp/}.}---just few of
them show their proofs.

One approach to address the proof-reconstruction problem is
justifying each inference rule of the \emph{source} system, the \ATP,
with a proof-assistant, the \emph{target} system. The target system
is the proof \emph{checker} in charge to verify the source system
reasoning.
These proof-assistants allow us to formalize the logical system used
in the proofs, \ie, logical constants, axioms, inference rules,
hypotheses, and theorems. A proof-reconstruction tool provides such
an integration, translating the derivation generated by the prover
into the formalism of the proof-assistant.

@@

We describe a formal reconstruction of proofs generated by the \Metis
prover~\cite{hurd2003first}---our source system---in Martin-Löf type
theory~\cite{Nordstrom-Petersson-Smith-1990}. We
formalize the subset of the \Metis inference rules for the
propositional logic fragment using a syntactical treatment.
The \Metis reasoning was formalized in
\Agda~\cite{agdateam}---our target system---in two
libraries~\cite{AgdaProp,AgdaMetis} and we implemented a
proof-reconstruction tool named \Athena~\cite{Athena} written in
\Haskell that is able to generate \Agda proof-terms for \Metis
derivations. During writing this document, our formalization helps to report two programming errors in \Metis.
A bug\footnote{Issue No.\,2 at \url{https://github.com/gilith/metis/issue/4}} in the printing of the derivation and a soundness bug\footnote{Issue No.\,4 at \url{https://github.com/gilith/metis/issue/4}} in the stripping of the goal.

>>>>

In order to give confidence to the \ATP users many of these systems
have started to include in their outputs the full derivations
associated to the proved theorems. However, existing research
recognizes that in many cases these derivations encode non-trivial
reasoning hard to reconstruct and therefore hard to
verify~\cite{paulson2007source,Keller2013}.

Proof-reconstruction addresses this problem. Since many \ATPs are
poor-documented, this problem becomes in mostly cases a reverse
engineering task to verify the prover reasoning. The usual is the
reconstruction is made by another and not by the developers of the
\ATP. Therefore, the presentation of the derivations generated by the
prover plays an important role in proof-reconstruction.

To verify such automatically generated derivation by the prover,
it is convenient to have them in a
consistent format, that is, a full script describing the derivation
step-by-step with  exhaustive details and without ambiguities. For
example, for classical propositional logic (henceforth \CPL) from a
list of at least forty \ATPs---available from the Web service
\name{SystemOnTPTP} of the TPTP World\footnote{\url{http://www.cs.miami.edu/~tptp/}.}---just few of
them show their proofs.

One approach to address the proof-reconstruction problem is
proving each deduction of the prover, the \emph{source} system,
with a formalization of the prover reasoning in a proof-assistant, the \emph{target} system. The target system is the
proof \emph{checker} in charge to verify the source system reasoning
for each derivation.
These proof-assistants allow us to formalize the logical system used
in the proofs, \ie, logical constants, axioms, inference rules,
hypotheses, and theorems.
A proof-reconstruction tool provides such an
integration, translating the derivation generated by the prover into
the formalism of the proof-assistant.

@@

We describe a formal reconstruction of proofs generated by the \Metis
prover~\cite{hurd2003first}---our source system---in Martin-Löf type
theory~\cite{Nordstrom-Petersson-Smith-1990}. We
formalize the subset of the \Metis inference rules for the
propositional logic fragment using a syntactical treatment.
The \Metis reasoning was formalized in
\Agda~\cite{agdateam}---our target system---in two
libraries~\cite{AgdaProp,AgdaMetis} and we implemented a
proof-reconstruction tool named \Athena~\cite{Athena} written in
\Haskell that is able to generate \Agda proof-terms for \Metis
derivations. During writing this document, our formalization helps to report two programming errors in \Metis.
A bug\footnote{Issue No.\,2 at \url{https://github.com/gilith/metis/issues/2}} in the printing of the derivation and a soundness bug\footnote{Issue No.\,4 at \url{https://github.com/gilith/metis/issues/4}} in the stripping of the goal.

### 2017-11-23 paper/sections/proof-reconstruction.tex
First version of the document - November 22, 2017 (#51)
* [ paper ] two first paragraphs in the introduction.

* [ bib] Fixed warning.

(Warning--empty publisher in Bertot2004).

* [ bib ] Fixed warning.

(Warning--empty booktitle in Bove2005).

* [ bib ] Fixed warning.

(Warning--empty booktitle in Een2004).

* [ bib ] Fixed warning.

(Warning--empty journal in Altenkirch2015).

* [ bib ] Fixed warning.

(Warning--empty year in deNivelle2003).

* [ paper ] add the third paragraph in the introduction.

* [ paper ] added the fourth paragraph in the introduction.

* [ paper
] fixed some text in

We have finished the formalization of every inference rule in
a \Metis proof, we are able to justify step-by-step any proof
for a problem in propositional logic.
For instance, we tested successfully the translation by \Athena
jointly with the \Agda formalizations of the rules mentioned above
with more than eighty representative theorems in propositional
logic. An interested reader can test the
problems~\cite{Prieto-Cubides2017} in the \Athena tool
repository~\cite{Athena}.

>>>>

We have finished the formalization of every inference rule in
a \Metis derivation, we are able to justify step-by-step any proof
for a problem in propositional logic.
For instance, we tested successfully the translation by \Athena
jointly with the \Agda formalizations of the rules mentioned above
with more than eighty representative theorems in propositional
logic. An interested reader can test the
problems~\cite{Prieto-Cubides2017} in the \Athena tool
repository~\cite{Athena}.

### 2017-11-23 paper/sections/proof-reconstruction/strip.tex
First version of the document - November 22, 2017 (#51)
* [ paper ] two first paragraphs in the introduction.

* [ bib] Fixed warning.

(Warning--empty publisher in Bertot2004).

* [ bib ] Fixed warning.

(Warning--empty booktitle in Bove2005).

* [ bib ] Fixed warning.

(Warning--empty booktitle in Een2004).

* [ bib ] Fixed warning.

(Warning--empty journal in Altenkirch2015).

* [ bib ] Fixed warning.

(Warning--empty year in deNivelle2003).

* [ paper ] add the third paragraph in the introduction.

* [ paper ] added the fourth paragraph in the introduction.

* [ paper
] fixed some text in

The $\fstrip₀$ function yields the conjunction of subgoals that implies the
goal of the problem in the \Metis \TSTP derivations but is not a structurally
recursive function.  Instead, we present its structural recursive version
in~\eqref{eq:strip-fixed}, the $\fstrip₁$ function after applying the bounded
technique described in Section~\ref{ssec:structural-recursion}. We define the
$\fstrip₁$ function based on the definition from the \Metis source code.

@@

\begin{verbatim}
  fof(goal, conjecture, (p ∧ r ∧ q)).
  fof(s₁, p, inf(strip, goal)).
  fof(s₂, p ⇒ r, inf(strip, goal)).
  fof(s₃, (p ∧ r) ⇒ q, inf(strip, goal)).
  fof(neg₁, ¬ p, inf(negate, s₁)).
  ...
  fof(neg₂, ¬ (p ⇒ r), inf(negate, s₂)).
  ...
  fof(neg₃, ¬ ((p ∧ r) ⇒ q), inf(negate, s₃)).
\end{verbatim}
\end{myexamplenum}

>>>>

The $\fstrip₀$ function defined in~\eqref{eq:strip-zero} yields the
conjunction of subgoals that implies the goal of the problem in the
\Metis \TSTP derivations. Nonetheless, this function is not a
structurally recursive function (see more details in
Appendix~\ref{app:strip-cm}).
Therefore, we present the $\fstrip₁$ function
in~\eqref{eq:strip-fixed} as the structurally recursive version
of the $\fstrip₀$ function by applying the
bounded technique described in
Section~\ref{ssec:structural-recursion}.
We define the $\fstrip₁$ function based on the reading of the \Metis source code.

@@

\begin{verbatim}
  fof(goal, conjecture, p ∧ r ∧ q).
  fof(s₁, p, inf(strip, goal)).
  fof(s₂, p ⇒ r, inf(strip, goal)).
  fof(s₃, (p ∧ r) ⇒ q, inf(strip, goal)).
  fof(neg₁, ¬ p, inf(negate, s₁)).
  ...
  fof(neg₂, ¬ (p ⇒ r), inf(negate, s₂)).
  ...
  fof(neg₃, ¬ ((p ∧ r) ⇒ q), inf(negate, s₃)).
\end{verbatim}
\end{myexamplenum}

### 2017-11-23 paper/sections/type-theory.tex
First version of the document - November 22, 2017 (#51)
* [ paper ] two first paragraphs in the introduction.

* [ bib] Fixed warning.

(Warning--empty publisher in Bertot2004).

* [ bib ] Fixed warning.

(Warning--empty booktitle in Bove2005).

* [ bib ] Fixed warning.

(Warning--empty booktitle in Een2004).

* [ bib ] Fixed warning.

(Warning--empty journal in Altenkirch2015).

* [ bib ] Fixed warning.

(Warning--empty year in deNivelle2003).

* [ paper ] add the third paragraph in the introduction.

* [ paper ] added the fourth paragraph in the introduction.

* [ paper
] fixed some text in

By following the Curry-Howard correspondence (see, for
example,~\cite{Wadler-2015}), a formula corresponds to a \emph{type},
and one proof of that formula is a \emph{term} of the correspondent type.
Therefore, inhabitants types are such formulas with proofs, that is, theorems.

@@

We produce derivation trees inside the judgments using the derivation
rules. Since type theory can be seen as typed $λ$-calculus with
dependent function types, evaluation of $λ$-terms also called
\emph{normalization} of the proof-terms, are reductions using the
system inference rules.

Therefore, the proof verification task becomes in type theory as
verifying that the proof-term has the correspondent type of the
theorem. We know this process as \emph{type-checking}.
This feature of type theory allow us to verify a proof generated by an \ATP by reconstructing its proof-term to type-check the proof. For such a
purpose, we use a proof assistant based on type theory like \Agda to
delegate this task. % However, this approach has some weak points
discussed in~\cite{Bezem2002}. Some limitations from the type theory
point of view for proof-reconstruction is described in the following
sections.

>>>>

By following the Curry-Howard correspondence (see, for
example,~\cite{Wadler-2015}), a formula corresponds to a \emph{type},
and one proof of that formula is a \emph{term} of the correspondent
type. Therefore, inhabitants types are such formulas with proofs, they are theorems.

@@

We produce derivation trees inside the judgments using the derivation
rules. Since type theory can be seen as typed $λ$-calculus with
dependent function types, evaluation of $λ$-terms also called
\emph{normalization} of the proof-terms, is a process of reductions with the system inference rules.

Therefore, the proof verification task becomes in type theory as
verifying that the proof-term has the correspondent type of the
theorem. We know this process as \emph{type-checking}.
This feature of type theory allow us to verify a proof generated by an \ATP by reconstructing its proof-term to type-check the proof. For such a
purpose, we use a proof-assistant based on type theory like \Agda to
delegate this task. % However, this approach has some weak points
discussed in~\cite{Bezem2002}. Some limitations from the type theory
point of view for proof-reconstruction is described in the following
sections.
